---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# TableCleaneR

<!-- badges: start -->
<!-- badges: end -->

TableCleaneR provides small, focused helpers to **collapse duplicate rows defined by key columns** in tibbles/data frames, in a way that is safe for downstream analysis:

- Key combinations are made unique by **collapsing** rows sharing the same keys.
- Selected columns can be **concatenated** to preserve information (e.g. sources, comments).
- Other divergent non-key fields are **replaced with typed `NA`** to avoid silently mixing inconsistent values.
- Companion functions help **audit** which groups require concatenation and subset collapsed outputs accordingly.

## Installation

### Development version (GitHub)

```r
# install.packages("pak")
pak::pak("besibo/TableCleaneR")
```

(Alternatively)

```r
# install.packages("devtools")
devtools::install_github("besibo/TableCleaneR")
```

## Core functions

TableCleaneR currently exposes four functions:

- `collapse_by_keys()`: collapse duplicates to one row per key group; concatenate `.concat` columns when needed; replace other divergent columns with `NA`.
- `collapse_candidates()`: return the **original input rows** belonging to groups where concatenation would be needed for `.concat`.
- `collapse_concat_only()`: return the **collapsed output** restricted to groups where concatenation actually occurred.
- `collapse_no_concat()`: return the **collapsed output** restricted to groups where no concatenation was needed for `.concat`.

## Quick start

```{r}
library(TableCleaneR)
library(dplyr)
library(tibble)

df <- tibble(
  exam.num_collec = c(1, 1, 1, 2, 2),
  mat.matrice = c("SER", "SER", "SER", "PLAS", "PLAS"),
  spe.denomination = c(
    "E. coli",
    "E. coli",
    "E. coli",
    "S. aureus",
    "S. aureus"
  ),
  result = c("POS", "NEG", "POS", "NEG", "NEG"), # divergent (non-concat) for key=1
  value = c(10, 10, 12, 5, 5), # divergent (non-concat) for key=1
  commentaire = c("first", NA, "repeat", "ok", "ok"), # concat target
  source_info = c("labA", "labA", "labB", "labC", NA), # concat target
  flag = c(TRUE, TRUE, TRUE, FALSE, FALSE) # constant within groups
)

df
```

### 1) Collapse duplicates safely

```{r}
collapsed <- df %>%
  collapse_by_keys(
    exam.num_collec,
    mat.matrice,
    spe.denomination,
    .concat = c(commentaire, source_info),
    sep = " | ",
    na_rm = TRUE,
    warn = TRUE
  )

collapsed
```

Expected behaviour (for the key group `(1, "SER", "E. coli")`):

- `commentaire` becomes `"first | repeat"` (distinct non-NA values concatenated),
- `source_info` becomes `"labA | labB"`,
- `result` and `value` become `NA` (divergent and not selected for concatenation),
- `flag` stays `TRUE` (constant within the group).

### 2) Audit: which original rows would require concatenation?

```{r}
candidates <- df %>%
  collapse_candidates(
    exam.num_collec,
    mat.matrice,
    spe.denomination,
    .concat = c(commentaire, source_info),
    na_rm = TRUE
  )

candidates
```

This returns the **input rows** for key groups where at least one `.concat` column has more than one distinct value.

### 3) Keep only collapsed groups where concatenation occurred

```{r}
only_concat <- df %>%
  collapse_concat_only(
    exam.num_collec,
    mat.matrice,
    spe.denomination,
    .concat = c(commentaire, source_info),
    sep = " | ",
    na_rm = TRUE,
    warn = FALSE
  )

only_concat
```

### 4) Keep only collapsed groups where no concatenation was needed

```{r}
no_concat <- df %>%
  collapse_no_concat(
    exam.num_collec,
    mat.matrice,
    spe.denomination,
    .concat = c(commentaire, source_info),
    na_rm = TRUE,
    warn = FALSE
  )

no_concat
```

## Tidyselect support

Column selection for `.concat` uses **tidyselect** (e.g. `starts_with()`, `contains()`, `matches()`, `where()`):

```{r}
df %>%
  collapse_by_keys(
    exam.num_collec,
    mat.matrice,
    spe.denomination,
    .concat = starts_with("source"),
    sep = " | "
  )
```

## Notes

- `.concat` columns return a character result when concatenation is needed.
- For non-`.concat` columns, divergent groups are replaced by typed `NA` (to reduce accidental mixing of inconsistent values).
- All functions are designed for use in tidyverse pipelines; selection arguments are captured using tidyeval/tidyselect (no premature evaluation).
